---
title: "math modelling"
author: "Rose Masinde"
date: "2024-05-27"
output: html_document
---

```{r}
library(deSolve)
library(tidyverse)
library(DiagrammeR)
```

1.Generate the schematic framework to illustrate the compartments and the rates across different compartments

#succeptible =population which are not infected but are at chance of getting infected #infected=population infected and can transmit covid #recovered=population recovered and are immune

#transmition rate=beta,rate at which the succeptible get infected with covid

#recovery rate=gamma,rate at which the infected recover from the disease #simulation period is the follow up period.

2 initialize the parameters and initial state values\

```{r}
N<-50000000
Initial_Infected_Individuals<-500
Initial_Recovered_Individuals<-0
Duration_of_Infectiousness<-10 
Transmission_Rate<-0.3 
Recovery_Rate<-1/10 
Simulation_Period<-180

 
initial_susceptible <-N- Initial_Infected_Individuals - Initial_Recovered_Individuals
initial_susceptible
```

Initialize state values

```{r}

initial_state_values <- c(S = initial_susceptible, I = Initial_Infected_Individuals, R = Initial_Recovered_Individuals)
initial_state_values
parameters <- c(beta = Transmission_Rate, gamma = Recovery_Rate)
times <- seq(from = 0, to = Simulation_Period, by = 1)
times
```

# QUESTION 2 #Implement the SIR model in R using the parameters defined in the scenario.

#Your implementation should define the model equations and solve them over a period of 180 days. #Explain the role of each parameter in the SIR model and how it influences the dynamics of the epidemic.

```{r}
sir_model <- function(time, state, parameters) {     
  with(as.list(c(state, parameters)), { 
    # compute derivatives
    dS<- (-beta*S*I)
    dI <- (beta*S*I) - (gamma * I)
    dR <- (gamma * I)
    return(list(c(dS, dI, dR)))                             
  })
}
```

#solve the sir model

```{r}
output <- as.data.frame(ode(y = initial_state_values, 
                            times = times, 
                            func = sir_model,
                            parms = parameters))
output
```

#Explain the role of each parameter in the SIR model and how it influences the dynamics of the epidemic. #beta is the rate at which the succeptible can get infected,transmission rate is Average number of individuals that one infected individual can infect per day #gamma is the rate at which the infected can recover,the recovery rate are the iknfected individuals who recover per day

#QUESTION 3 #Create a line plot showing the number of susceptible, infected, recovered, and vaccinated individuals over time for the model implemented. #data analysis between the different states

```{r}
dat <- output %>% pivot_longer(cols = c(S,I,R),
                               names_to = "state",
                               values_to = "value")
knitr::kable(head(dat,10))
```

#data visualization between different states

```{r}
dat %>% ggplot(aes(x = time,
                   y = value,
                   color = state,
                   
                   
                   group = state))+
  geom_line()+
  xlab("Time(Days)")+
  ylab("Number of persons")+
  labs(title = "model of Covid in kenya")
```

\#**QUESTION 4 #Assume a vaccination of 20% of the population. Discuss how the vaccination rate affects the epidemic curve**

#ANS a vaccination will increase the average number of individuals who will be resistant to the infection.Therefore the infected rate will decrease and the recovery rate will increase.This is because of the resistance to covid therefore covid infection will reduce.

**#What changes do you observe in the number of infected individuals when the vaccination rate is doubled**? #ANS using the plots the curves will be more flatter nearly the same especially the succeptible and the recovery.This is due to that many in the population will be resistant and a few will be infected and a few will be recovered.So covid will be resisted

**#Explain the importance of the transmission rate (𝛽) and how public health interventions could reduce this rate** #Explains how the disease is infected from from the infected individuals to the susceptible. #pub;ic health will reduce by introducing vaccinations to the individuals,Through lockdown and curfews to reduce the contact among people.Wear masks to prevent the infection of covid.Wash hands and practice social distancing.
