---
title: "stochastic process"
author: "Rose Masinde"
date: "2024-05-27"
output: html_document
---

\#**QUESTION 1: #An independent landscape contractor works on a weekly basis. Each week he works (W), there is a #probability of 80% that will be called again to work the following week. Each week he is not working \#(N), there is a probability of only 60% that he will be called again to work. #Find #a) Define the transition probability matrix**

```{r}
transition  <- matrix(c(0.8, 0.2, 0.6, 0.4), nrow = 2, byrow = TRUE)
transition
```

Name the rows and columns for clarity

```{r}
rownames(transition ) <- c("W", "N")
colnames(transition) <- c("W", "N")
transition

```

# 

# transition probability diagram

```{r}
library(DiagrammeR)

diagram <- "
digraph {
  W [label='W ', shape=circle]
  N [label='N ', shape=circle]
  
  W -> W [label='0.8']
  W -> N [label='0.2']
  N -> W [label='0.6']
  N -> N [label='0.4']
}
"

grViz(diagram)

```

#The probability that if he is working now, then he will be working in 2 weeks \# the square of the transition matrix

```{r}
transition_2 <- transition %*% transition
transition_2

probability_ww <- transition_2[1, 1]
probability_ww  


```

#The probability that if he is not working now, then he will be working in 2 weeks

```{r}
probability_nw<-transition_2[2,1]
probability_nw

```

#The probability that if he is working now, then he will not be working in 2 weeks

```{r}
probability_wn<-transition_2[1,2]
probability_wn
```

The probability that if he is not working now, then he will not be working in 2 weeks

```{r}
probability_nn<-transition_2[2,2]
probability_nn
```

#QUESTION 2 #A market analyst is interested in whether consumers prefer Dell or Gateway computers. Two market #surveys taken one year apart reveals the following: \# − 10% of Dell owners had switched to Gateway and the rest continued with Dell. #− 35% of Gateway owners had switched to Dell and the rest continued with Gateway. #At the time of the first market survey, 40% of consumers had Dell computers and 60% had Gateway. #a) Define the transition probability matrix

```{r}
transition2  <- matrix(c(0.9, 0.1, 0.35, 0.65), nrow = 2, byrow = TRUE)
transition2
```

Name the rows and columns for clarity

```{r}
rownames(transition2 ) <- c("D", "G")
colnames(transition2) <- c("D", "G")

transition2
```

transition diagram

```{r}
library(DiagrammeR)

diagram <- "
digraph {
  W [label='D', shape=circle]
  N [label='G', shape=circle]
  
  W -> W [label='0.9']
  W -> N [label='0.1']
  N -> W [label='0.35']
  N -> N [label='0.65']
}
"

grViz(diagram)

```

#What percentage will buy their next computer from Dell \# Initial state vector

```{r}
initial_state <- c(D = 0.4, G = 0.6)
```

# Compute the next state

```{r}
next_state <- initial_state %*% transition2
next_statepercent<-(next_state[1,1]*100)
next_statepercent
```

#What percentage will buy their second computer from Dell

```{r}
second_state <- next_state %*% transition2
second_statepercent<-(second_state[1,1]*100)
second_statepercent
```

#Suppose that each consumer buys a new computer each year, what will be the market #distribution after 4 year

```{r}
after4years <- initial_state %*% (transition2 %*% transition2 %*% transition2 %*% transition2)
print(after4years)
```
