---
title: "statistical modelling"
author: "Rose Masinde"
date: "2024-09-11"
output: html_document
---

UPLOADING THE EXCEL FILE

```{r}
library(readxl)
data<-read_xlsx("C:/Users/user/Downloads/Student health survey data.xlsx")
data
```

Multiple linear regression in R

```{r}
model<-lm(incmnth~age+sex+homecost+mobilecost+transport+food+entertain+veg+food+medication+height+weight,data=data)
summary(model)
```

Assumptions:linear relationship

```{r}
# Basic scatter plot
#plot(x, y, main="Scatter Plot Example", xlab="X Axis", ylab="Y Axis", pch=19, col="blue")
library(ggplot2)
library(tidyverse)
ggplot(data) +
geom_point(aes(x = food, y = incmnth))
# Plot the matrices between
# 4 variables giving 12 plots.
# One variable with 3 others
# and total 4 variables.
pairs(~age+sex+homecost+mobilecost+transport+food+entertain+veg+food+medication+height+weight, data = data,
main = "Scatterplot Matrix")
```

income has a linear relationship with food but not with age

```{r}
library(tidyverse)
library(ggplot2)
ggplot(data, aes(x = age, y = incmnth)) +
  geom_point() +
  labs(title = "Scatter Plot of Age vs. Income", x = "Age", y = "Income") +
  theme_minimal()
```

Assumption 2:No multicollinearity

```{r}
library(car)
summary(model)
vif(model)
```

A value of 1 vif indicates there is no correlation between a given predictor variable and any other predictor variables between 1 and 5 indicates moderate correlation

Assumption 3:independence

```{r}
library(car)
#perform Durbin-Watson test
durbinWatsonTest(model)
```

LASSO REGRESSION

```{r}
#define response variable
y <- mtcars$hp

#define matrix of predictor variables
x <- data.matrix(mtcars[, c('mpg', 'wt', 'drat', 'qsec')])


```

```{r}
library(glmnet)

#perform k-fold cross-validation to find optimal lambda value
cv_model <- cv.glmnet(x, y, alpha = 1)

#find optimal lambda value that minimizes test MSE
best_lambda <- cv_model$lambda.min
best_lambda

#[1] 5.616345

#produce plot of test MSE by lambda value
plot(cv_model) 

```

```{r}
install.packages("remotes")
```
